---
title: "Benchmarking future.apply"
date: 2026-03-25
weight: 10
summary: "A very brief introduction on using the DataFrames package in Julia"
tags: ["R", "programming", "loops", "parallelization"]
output: hugodown::md_document
---

```{r setup, include=FALSE}
library(future.apply)
library(microbenchmark)
library(ggplot2)
library(knitr)

# Setup parallel workers
plan(multisession, workers = parallel::detectCores() - 1)
```

## The Three Contenders
1. **Standard `for` loop**: Manual iteration (pre-allocated).
2. **`lapply`**: The functional, sequential R standard.
3. **`future_lapply`**: The parallelized version.

---

## Experiment 1: The "Cheap" Task
In this scenario, we do something very fast: calculating the mean of 1,000 numbers. 

```{r cheap_benchmark}
n <- 200
data_list <- replicate(n, rnorm(1000), simplify = FALSE)

bench_cheap <- microbenchmark(
  for_loop = {
    res_for <- vector("list", n)
    for(i in 1:n) res_for[[i]] <- mean(data_list[[i]])
  },
  standard_apply = lapply(data_list, mean),
  future_apply   = future_lapply(data_list, mean),
  times = 10
)

# Generate Table
kable(summary(bench_cheap), caption = "Cheap Task Results (milliseconds)")

# Generate Figure
autoplot(bench_cheap) + labs(title = "Cheap Task: Parallel Overhead is Visible")
```

---

## Experiment 2: The "Expensive" Task
In this scenario, we simulate "heavy" work by adding a tiny delay (`Sys.sleep`). This mimics complex statistical modeling or web scraping.

```{r expensive_benchmark}
n_heavy <- 20
data_heavy <- replicate(n_heavy, rnorm(10), simplify = FALSE)

# A function that takes 0.1 seconds per call
heavy_func <- function(x) {
  Sys.sleep(0.1)
  mean(x)
}

bench_expensive <- microbenchmark(
  for_loop = {
    res_for <- vector("list", n_heavy)
    for(i in 1:n_heavy) res_for[[i]] <- heavy_func(data_heavy[[i]])
  },
  standard_apply = lapply(data_heavy, heavy_func),
  future_apply   = future_lapply(data_heavy, heavy_func),
  times = 2 # Low iterations because it's slow!
)

# Generate Table
kable(summary(bench_expensive), caption = "Expensive Task Results (seconds)")

# Generate Figure
autoplot(bench_expensive) + labs(title = "Expensive Task: Future Wins Big")
```
